---
title: "Using the DMR Scan Package"
author: "Christian M Page"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{DMRScan Vignette}
  %\VignetteEngine{knitr::rmarkdown}
---

# Abstract
###Motivation
DNA methylation plays an important role in human health and disease, and methods 
for the identification of differently methylated regions are of increasing interest. 
There is currently a lack of statistical methods which properly address 
multiple testing, i.e. genome-wide significance for differentially methylated regions.

###Methods
We introduce a scan statistic (DMRScan), which overcomes these limitations, but 
don't suffers from the same limitations as alternative R packages, such as 
"Bumphunter" and "DMRCate".

### Data Set
In this package, we have included a small example data set, which is an extract 
of chromosome 22 from a large whole genome RRBS study. 

# Work flow and use of DMRScan
## Data inputs

### Example data
Start by loading the package and have the methylation data ready in the workspace. 
The minimal data requirements for this package to run, is a list of numerics, 
with sequentially ordered test statistics, on which the sliding window is used. 
In this example, we will be using an example data set to generate the needed input. 

```R
library(DMRScan)
data(DMRScan.methylationData) ## Load methylation data from chromosome 22, with 52018 CpGs measured
data(DMRScan.phenotypes) ## Load phenotype (end-point for methylation data)
``` 

### Test statistics
> Note that the DMRScan do not require raw data to be used, but only test statistics. 
To generate test statistics for use in DMRScan, we use logistic regression on 
the example DNA methylation data, and the phenotype data as endpoint.
<<<<<<< HEAD
```R
observations <- apply(DMRScan.methylationData,1,function(x,y)summary(glm(y ~ x, family = binomial(link = "logit")))$coefficients[2,3], y = DMRScan.phenotypes)
head(observations)
=======
```[r}
test.statistics <- apply(DMRScan.methylationData,1,function(x,y)
                  summary(glm(y ~ x, family = binomial(link = "logit")))$coefficients[2,3], 
                  y = DMRScan.phenotypes)
head(test.statistics)
>>>>>>> c23fb8d915f4b124ad2c4c58590dfbecad9f03b6
```
This will return a sequence of test statistics; which is the data that is used 
further. The output which will be used further is given below:

| chr22.17061793 | chr22.17062497 | chr22.17062614 | chr22.17063057 | chr22.17063105 | chr22.17063107|
| ---------------|----------------|----------------|----------------|----------------|---------------|
|   0.02924113   |    0.65795471  |   0.86192751   |  -0.39548318   |   -0.73786743  | 0.28811812    |

The sliding window function in DMRScan is not dependent on raw methylation data, 
but rather the test statistic for each genomic position. This allows for much 
grater flexibility when applying different models to the raw data, and can also 
be used with meta-analysis values, and not only primary studies. 

### Clustering of the test statics into "Chunks"
To apply the sliding window on a set of test statistics, we cluster them into 
chunks with not to much space between the probes. The maximum allowed distance 
between two probes in the same region is given by `max.gap`, and the minimum 
number of probes within a cluster is given by `min.cpg`.

The clustering is done automatically by `make.cpg.cluster()` and requires four
inputs; 
 - A sequence of test statistics
 - The corresponding genomic position for the test statistics
 - The maximum allowed gap (in base pairs) inside a sliding window, 
        i.e. the maximum gap allowed in a cluster
 - The minimum number of probes in a cluster. This is set to 2 at default. 

To identify the coordinate to each test statistic given in the previous table, 
we split the names inherited from the methylation values into its chromosomal 
part, and the base pair location.
<<<<<<< HEAD
```R
pos <- matrix(as.integer(unlist(strsplit(names(observations),split="chr|[.]"))), ncol = 3, byrow = TRUE)[,-1] 
=======
```{r}
pos <- matrix(as.integer(unlist(strsplit(names(test.statistics),split="chr|[.]"))), 
                                                        ncol = 3, byrow = TRUE)[,-1] 
>>>>>>> c23fb8d915f4b124ad2c4c58590dfbecad9f03b6
head(pos)
```
The chromosome number and genomic position is here stored in the same object, a 
two column matrix. This is not needed, and if more convinient, the two peaces of
information can be stored in separate objects.

To generate the clusters, we also need to set the additional clustering 
parameters:

```R
## Minimum number of CpGs in a tested cluster 
min.cpg <- 3  

## Maxium distance (in base-pairs) within a cluster 
## before it is broken up into two seperate cluster
max.gap <- 750  

```
This, together with the test statistic vector from earlier, allows us to generate 
a list of clusters where the sliding window is operated. 

```R
regions <- make.cpg.regions(test_statistic = test.statistics, chr = pos[,1], pos = pos[,2], 
                                              max.gap = max.gap, min.cpg = min.cpg)
``` 
## DMRScan 
We are now ready to run the sliding window on the clusters. To run a sliding 
window, a few things beside the clusters are needed:
- The sizes of the windows. This can be either a single window size, or 
a sequence of windows to be tested.
  - If a window sequence is given, the multiple testing adjustment will take this
  into account, and have a more stringent threshold for different window sizes. 
- The window thresholds need to be calculated. This is dependent on a number of
potential settings from the used. 

### Calculating window thresholds
Three different approaches to estimate the window thresholds is given in the 
package. The different options are: 
- A full MCMC model with simulation of null model `model = "mcmc"`, with given 
correlation structure for the null data, provided by `arima.sim()`.
- An faster MCMC called "important sampling", where only a subset of the data
is sampled. This is implemented in `model = "zhang"`, and is up to 700 times 
faster than the MCMC model, and simulation studies indicates that these to 
options are comparable in performance.
- An analytical solution to calculate the window thresholds, based on Siegmund
et.al (2012). Given by the model `model = "siegmund"`. 
Since this is a closed form expression, it is much faster to 
calculate compared to the two other thresholds. However, this option tends to 
give more conservative thresholds, with somewhat lower power, but fewer false
positive windows. An additional consideration with this method, is that it 
assumes that the test statistics follows an Ornstein-Uhlenbeck process, which 
may not always be the case. 

To calculate the threshold(s), a few parameters needs to be set; 
 - The method by which to estimate the threshold (see section above).
 - The window sizes. Can be either a single window size, or a sequence of 
 window sizes. 
 - The total number of observations in the study.

Additionally, for the Monte Carlo options, the number of iterations need to be 
specified, as well as the correlation structure. We will illustrate using only 
important sampling (which has a fixed correlation structure on the form of AR(2)) 
and the closed form expression from Siegmund et.al. 

```R
window.sizes <- 3:7 ## Number of CpGs in the sliding windows (can be either a single number or a sequence)
n.CpG        <- nCpG(regions) ## Number of CpGs to be tested
## Estimate the window threshold, based on the number of CpGs and window sizes 
## using important sampling
window.thresholds.importancSampling <- estimateWindowThreshold(nProbe = n.CpG, windowSize = window.sizes, method = "sampling", mcmc = 10000)

## Estimating the window threshold using the closed form expression
window.thresholds.siegmund <- estimateWindowThreshold(nProbe = n.CpG, windowSize = window.sizes, method = "siegmund")
``` 
### Identifying Differentially Methylated Regions
We now have all the data and parameters needed to run the `dmr_scan()` function. 
First using a window threshold estimated with _importanc sampling_, we have
```R
dmrscan.results   <- dmrscan(observations = regions, windowSize = window.sizes, windowThreshold = window.thresholds.importancSampling)
## Number of CpGs in the sliding windows (can be either a single number or a sequence)
window.sizes <- 3:7 

## Number of CpGs to be tested
n.CpG        <- nrow(DMRScan.methylationData) 

## Estimate the window threshold, based on the number of CpGs and window sizes 
## using important sampling
window.thresholds.importantSampling <- estimate_t_grid(k_grid = window.sizes, 
                                          L = n.CpG, method = "sampling", mcmc = 10000)

## Estimating the window threshold using the closed form expression
window.thresholds.siegmund <- estimate_t_grid(k_grid = window.sizes, 
                                           L = n.CpG, method = "siegmund")
``` 
### Identifying Differentially Methylated Regions
We now have all the data and parameters needed to run the `dmr_scan()` function. 
First using a window threshold estimated with _important sampling_, we have
```{r}
dmrscan.results   <- dmr_scan(obs = regions, k_grid = window.sizes, 
                              t_grid = window.thresholds.importantSampling)
## Print the result
print(dmrscan.results)
```
This will give the following result, with four genome wide significant regions. Two DMRs containing 3 CpGs and two containing 5 CpGs. 

| Genomic Coord. | Start |  End  |Length | Bump |   p.val.emp |           Pos          |
|----------------|-------|-------|-------|-----:|-------------|-------------------------
| chr22.51002050 | 51043 | 51045 |     3 | 2539 | 1.056161e-05| chr22:51002050-51002059|
| chr22.23801586 | 10807 | 10809 |     3 | 573  | 2.112323e-05| chr22:23801586-23801597|
| chr22.43821030 | 36791 | 36795 |     5 | 1925 | 2.217939e-04| chr22:43821030-43821098|
| chr22.46762751 | 41946 | 41950 |     5 | 2158 | 2.693212e-04| chr22:46762751-46762782|


When using a more stringent cut-off, generated by using the "_siegmund_" option in `dmr_scan()`, we get no significant regions on the same data set. Exemplified by the syntax below:

```R
dmrscan.results   <- dmr_scan(obs = regions, k_grid = window.sizes, 
                                t_grid = window.thresholds.siegmund)
## Print the result
print(dmrscan.results)
```
Which will return an object of type 'NA', since no genome wide significant regions were found.

End of vignette
